---
title: "Suicide Rates Within the U.S"
author: "Surina Soumpholphakdy, Natty Samuel, and Kristen Do"
date: "11/09/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
data <- read.csv("https://raw.githubusercontent.com/info201a-au2022/project_group_6_section_ab/main/data/who_suicide_statistics.csv")
library("ggplot2")
library("dplyr")
library("knitr")
```

## What is this file for?

## Abstract
Our main question is how often does suicide occur within the United States? We are concerned with the fact that due to many changes within recent years, the rates of suicide have increased immensely. To address this concern, we plan to take a closer look at suicide rates to bring awareness to mental health.

## Keywords
**Suicide** death caused by injuring oneself with the intent to die.

**Mental health** a person’s condition with regard to their psychological and emotional well-being

**Social wellbeing** a sense of belonging to a community and making a contribution to society

## Introduction
Do you think it’s normal for more than 700,000 people to commit suicide a year? A sad truth about today’s day in age is that suicide is the fourth leading cause of death for people 15-29 years-old. The only way we can change this and protect each other is by talking about these hard topics and being there for one another. It is very important for people to understand how much of an impact this has on our society as a whole. Whenever someone does commit suicide there are many more people who will attempt. In order to make a change and get people the help they need we need to shed light on this and have those hard conversations so that one day there can be a major change.

## Problem Domain
Within the year 2020, the United States alone estimated that there were 1.20 million attempted suicides and about 46,000 actual deaths. According to the CDC’s  article “Facts About Suicide”, when someone does suicide and or attempts it, the mental health and well-being of friends, loved ones, co-workers, and the community is negatively affected. Many times these loved ones will also experience feelings of shock, anger, guilt, symptoms of depression or anxiety. The only way we can change this and protect each other is by talking about these hard topics and being there for one another. In 2020 alone the racial/ethnic group with the highest suicide rates were non-Hispanic American Indians, Alaska Natives, and non-Hispanic Whites. Also the rate for suicide rate for males during 2020 was four-times higher than the females rate. People of all genders, ages, and ethnicities can be at risk for suicide. It is very important for people to understand how much of an impact this has on our society as a whole. 
 
Where in the world do we see the highest suicide rates? This data set depicts suicide numbers of people all across the world, including there nationality, sex, age, and year. Our project will be structured in a way where we will display various statistics and data sets of suicide rates amongst both male and female. From this you will also be able to compare the population of living people and people who committed suicide. We can also look at the year and see during which time periods people would have a higher numbers of committing suicide. What we can gather from analyzing this data is to make educated guesses as to why some people committed and make improvements in our society today so that we can reduce the number of suicides across the world.  
 
Direct stakeholders in regards to the correlation between mental health and social media could include other students, teens, social media users, and etc. On the other hand, indirect stakeholders, and people who are impacted without use would be families of teenagers, friends, and etc. Their are many factors that inevitably lead up to someone committing suicide. For some people it may be depression, mental disorders, and or substance abuse. One cause of depression and mental disorder can be drawn from the negative impacts of social media on teenagers and young adults. According to S.Dixon article about “Teen perspectives on negative effects of social media in the U.S. 2018”,  studies show the leading reasons why teenagers ultimately feel as though social media has a negative impact on their lives. According to his statistics the leading causes for teen depression is caused by bullying/spreading rumors through the use of social media.  For others may come from chronic pain their whole life or family history of mental disorder or substance abuse. 
 
The possible harm that can be drawn from suicide is that it doesn’t only affect that one person after that person has passed all the people who hear and or know about that person are now all affected. According to the CDC’s  article “Facts About Suicide”, when someone does suicide and or attempts it, the mental health and well-being of friends, loved ones, co-workers, and the community is negatively affected. Many times these loved ones/close friends also experience feelings of shock, anger, guilt, symptoms of depression or anxiety. The benefit to discussing this topic is that awareness and attention is brought to this subject. 

**_Related work_**

[Social Media Harm To Children](https://www.seegerweiss.com/social-media-youth-harm-2?gclid=CjwKCAjw5P2aBhAlEiwAAdY7dG3Ggi491WXnjm_MjVyEGLBZcOhT3FsYdbq14NlY8djpXf02sBtmORoCA08QAvD_BwE)

[Social Media and Mental Health](https://www.helpguide.org/articles/mental-health/social-media-and-mental-health.htm)

[Social Media Link to Decline in Mental Health](https://mitsloan.mit.edu/ideas-made-to-matter/study-social-media-use-linked-to-decline-mental-health)

[Does Social Media Use Cause Depression?](https://childmind.org/article/is-social-media-use-causing-depression/)

## Research questions
**_What age group has the highest suicide rate?_**
The question, “What age group has the highest suicide rate?” is an important topic to cover because it is detrimental that our group takes a closer look into what age group is suffering the most from poor mental health. This question is motivated by the want to find a solution for a specific age group in order to lessen the number of suicides within this specific age groups, since it is the age that is most impacted.

**_Is suicide more common in the male or female population?_**
The question, “Is suicide more common in the male or female population?” is an important topic to cover because it is necessary to narrow down the population in which suicide has a greater occurance. This question is motivated by the want to know whether or not a certain female or male genetic trait is the reason for a increase amount of deaths by suicide. From here, we are able to do some research and test our theory based on the data analysis.

**_What is the trend of suicide rates over the past few years?_**
The question, “What is the trend of the suicide rates over the past few years?” is an important topic to cover because we need to highlight the fact that suicide is a growing issue within our community. This question is motivated by the want to know whether or not the events within recent years have triggered more poor mental health and deaths by suicide.

## The Dataset
Our dataset specifically focuses on long-time trends of suicide rates between different countries. Suicide cases has been a prevalent issue within countries worldwide; not excluding the U.S. There are many factors that could lead or motivate an individual to commit suicide. These factors can be both social or political. Mental health is an important issue to recognize. Recognizing these rates allows others to be aware of the trends that could be increasing and how we can make plans of actions to decrease these trends. Our problem domain is to specifically focus on suicide rates within the U.S; which leads us to question and compare the trends of suicide over the years and potentially what its leading factors. Our dataset provides us with important information, alongside suicide rates, the years, sex group, age range, and population of the country. In order for us to be able to answer our research questions, these variables of data provided from our dataset are essential in being able to compare trending rates. With these variables we can specifically determine if suicide rates are more prevalent in relation to certain age groups, genders, or country. Overall, our goal is to fulfill the purpose of our dataset by spreading awareness towards these issues.

| [WHO Suicide Statistics](https://www.kaggle.com/datasets/szamil/who-suicide-statistics?resource=download) | |Year | Age | Suicides_no | Population |
| -----------| ----------- | ----------- | ---------- | ----------| ---------- |
| Countries | 43776 | 43776 | 43776 | 43776 | 43776


**Who collected the data?**:
WHO (World Health Organization)

**When?**:
2018

**For what purpose?**:
To observe where the the world as a whole was heading and whether it is getting worse or better in terms of suicide trends.

**How was the data collection effort funded?**:
WHO is funded by two main sources; Accessed contribution (AC) and voluntary contributions. Accessed contributions are made by countries who are members of the organization. While, voluntary contributions are made by regular members, philanthropist foundations, etc.

**Who is likely to benefit from the data or make money?**:
The creators, WHO, are more likely to benefit/make money from the data. This is because they are directly taking the donations to fund their collections. All though in hand, by funding these data collections, users are able to obtain information. Though, directly, WHO is being benefitted.

**How was the data validated and held secure?**:
They World Health Organization validate their data by collecting data from population-based sources such as, house-hold surveys, etc. and institution-based sources such as, health facilities, etc.

**Is it credible and trustworthy?**:
Yes this dataset is both credible and trustworthy. This dataset comes from a health research organization that focuses on health in general. The credibility can also be validated by their usage of multiple sources and expansion to different aspects of data that does not restrict them from potential limitations.

**How did you obtain the data?**:
We obtained the data by searching through the Kaggle database for data regarding suicide rates.

**Do you credit the source of the data?**:
Yes

## Expected Implications
Our expected implications regarding our research questions, could result in the idea of the fault lines that are created by policymakers and lack of action from social workers. There are a numerous factors that could motivate one to commit suicide. However, to commit these actions is also a reflection of the lack of action of others. The main motivation for suicide could be due to feeling neglected or downturned from others in times of need. This includes social workers, whether its a teacher, counselor, etc. And policymakers may also be a leading factor to these faults, as the lack of restrictive policies may give a leeway for those to freely commit harmful or negative actions towards someone without repercussions.

## Limitations
There are a few limitations when exploring our research for this project. The first main issue and limitation that our group could possibly come across would be the inability to find data on suicide rates. For example, we could use our knowledge of recent world events to justify the prediction of increased suicide rates, though it may be difficult to find data to support this theory. Another possible limitation we could come across throughout this research project would be our lack of experience with coding and creating websites. This is our first time coding for the majority of us and we expect that we will have difficulty coding the actual project or not being able to add features that we desire. 

## References

Miller , C. (2022, April 14). _Does social media use cause depression?_ Child Mind Institute. Retrieved October 31, 2022, from https://childmind.org/article/is-social-media-use-causing-depression/

Szamil. (2018, August 29). _Who suicide statistics._ Kaggle. Retrieved November 10, 2022, from https://www.kaggle.com/datasets/szamil/who-suicide-statistics?resource=download 

_Social Media and Mental Health._ HelpGuide.org. (n.d.). Retrieved October 31, 2022, from https://www.helpguide.org/articles/mental-health/social-media-and-mental-health.htm

_Social Media Harm to children._ Social Media Youth Harm Attorneys - Seeger Weiss LLP. (n.d.). Retrieved October 31, 2022, from https://www.seegerweiss.com/social-media-youth-harm-2?gclid=CjwKCAjw5P2aBhAlEiwAAdY7dG3Ggi491WXnjm_MjVyEGLBZcOhT3FsYdbq14NlY8djpXf02sBtmORoCA08QAvD_BwE

Walsh, D. (2022, September 14). _Study: Social media use linked to decline in mental health._ MIT Sloan. Retrieved October 31, 2022, from https://mitsloan.mit.edu/ideas-made-to-matter/study-social-media-use-linked-to-decline-mental-health
>>>>>>> c933bb08337f7e22b36e543fb7be9b79924d8158

## Dynamic Paragraph

After narrowing down the information found within our dataset, our group was able to focus on data collected from the United States. I was able to conclude values 1 and 2 from chart 1. **Value 1** consists of 3 reported deaths by suicide in 1980. In addition, **Value 2** consists of 9 reported deaths by suicide in 2009. From values 1 and 2, it is clear that the number of suicides increased over the years. From chart 2, we were able to find value 3. **Value 3** consists of the fact that from 1998 to 2016, the population is above 4e+07. From this value, it is clear that the population size increased. From chart 3, we were able to pull values 4 and 5. **Value 4** consists of 1 suicide by a female in 2008. **Value 5** consists of 3 suicides by males in 2008. From these values, we can conclude that more males die by suicide than females.

## Summary information
```{r summary}
```

## Table 
```{r Table}
source("../source/table.R")
table.R()
```
This aggregated table is a downsized and clean version of our who_suicide_statistics dataset. 
The purpose of downsizing our dataset is to steer more towards the focus of our research questions. 
We included information regarding data specifically from the U.S only; year, sex, both data from genders (female and male), age, suicide numbers, and population size. 

## Charts
```{r chart 1, echo=FALSE}
source("../source/scatterplot.R")
scatterplot.R()
```
I included the scatter plot containing the number of reported suicides per year based on the dataset because our group wanted to justify our prediction in regards to the increased suicide rates over the years. The information found within the chart reveals that over the years, the number of suicides have increased.

```{r chart 3, echo=FALSE}
source("../source/barchart.R")
barchart.R()
```
This is a bar chart that examines suicide numbers between genders over the years.
Within this chart, we include suicide numbers (y) over the years 1980 till 2010+ (x).
We also include two variables (female and male), depicted by the blue and orange bars shown on our chart. 
This is to show the drastic difference between suicides between genders. 



