---
title: "Suicide Rates Within the U.S"
author: "Surina Soumpholphakdy, Natty Samuel, and Kristen Do"
date: "11/09/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
data <- read.csv("https://raw.githubusercontent.com/info201a-au2022/project_group_6_section_ab/main/data/who_suicide_statistics.csv")
library("ggplot2")
library("dplyr")
library("knitr")
```

## What is this file for?





## Summary information
```{r summary}
```

## Table 
```{r Table}
source("../source/table.R")
table.R()
```
This aggregated table is a downsized and clean version of our who_suicide_statistics dataset. 
The purpose of downsizing our dataset is to steer more towards the focus of our research questions. 
We included information regarding data specifically from the U.S only; year, sex, both data from genders (female and male), age, suicide numbers, and population size. 

## Charts
```{r chart 3, echo=FALSE}
source("../source/barchart.R")
barchart.R()
```
This is a bar chart that examines suicide numbers between genders over the years.
Within this chart, we include suicide numbers (y) over the years 1980 till 2010+ (x).
We also include two variables (female and male), depicted by the blue and orange bars shown on our chart. 
This is to show the drastic difference between suicides between genders. 


